---
title: "Norman 2019"
format: html
editor: visual
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(zellkonverter)
  library(SingleCellExperiment)
  library(limma)
  library(Matrix)
  library(dplyr)
  library(purrr)
  library(tibble)
  library(ggplot2)
  library(posterior)
  library(bridgesampling)
  library(rstan)
  library(io)
})

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

```{r}
# Load data
norman <- readH5AD("/Users/zhangrui/Desktop/precmat/Norman_2019.h5ad",
                   reader = "R", use_hdf5 = TRUE)

# Keep control + single-gene perturbations
perturb <- as.character(colData(norman)$perturbation_name)
single <- !grepl("\\+", perturb)
sce_single <- norman[, perturb == "control" | single]

# Use raw counts
C <- assay(sce_single, "counts")

# Library size
lib_size <- Matrix::colSums(C)

# Remove cells with zero library size
keep_cells <- lib_size > 0
sce_single <- sce_single[, keep_cells]
C <- assay(sce_single, "counts")
lib_size <- Matrix::colSums(C)

ptb_raw <- as.character(colData(sce_single)$perturbation_name)

genes <- rownames(sce_single)
perts <- setdiff(sort(unique(ptb_raw)), "control")

# Offset for log(counts/lib_size + offset)
offset <- 1

cat("Number of genes:", length(genes), "\n")
cat("Number of perturbations:", length(perts), "\n")
```

```{r, echo = FALSE, message = FALSE, warning = FALSE, include = FALSE}
# Run differential expression analysis via limma for each perturbation
run_limma_one_perturbation <- function(g) {
  message("Running limma for perturbation: ", g)

  # Find cells belonging to either the control or the current perturbation
  idx <- which(ptb_raw %in% c("control", g))

  # Keep only the control and perturbation cells
  C_sub <- as.matrix(C[, idx, drop = FALSE])

  # Keep only library sizes corresponding to these selected cells
  lib_sub <- lib_size[idx]

  # Log normalization: divide raw umi by the library size and adding an offset
  E <- log(sweep(C_sub, 2, lib_sub, "/") + offset)

  # Control is the reference
  group <- factor(ifelse(ptb_raw[idx] == g, "perturb", "control"),
                  levels = c("control", "perturb"))

  # Construct design matrix
  design <- model.matrix(~ group)

  # Fit gene-wise linear model using limma to estimate coefficients
  fit <- lmFit(E, design)

  # Apply empirical Bayes moderation
  fit <- eBayes(fit, robust = TRUE)

  # Get BH adjusted p-values (q-value) by retaining all genes without sorting
  tt <- topTable(fit, coef = "groupperturb", number = Inf,
                 sort.by = "none", adjust.method = "BH")

  tibble(
    perturbation = g,
    gene = rownames(tt),
    logFC = tt$logFC,
    B = tt$B,
    p_value = tt$P.Value,
    q_value = tt$adj.P.Val
  )
}

# all_de <- bind_rows(lapply(perts, run_limma_one_perturbation))
# saveRDS(all_de, "/Users/zhangrui/norman_limma_all_perturbations.rds")
all_de <- readRDS("/Users/zhangrui/norman_limma_all_perturbations.rds")
```

```{r}
# Convert B to posterior probability of differential expression
all_de <- all_de %>% mutate(posterior_prob = plogis(B))

# Construct a binary gene regulatory network and calculate degree distribution
make_edges_and_degrees <- function(all_de, posterior_cutoff, genes, perts,
                                   remove_self_edges = FALSE) {

  # Keep perturbation -> gene effects with posterior probability above cutoff
  edge_df <- all_de %>% filter(posterior_prob >= posterior_cutoff)

  # Remove self-regulatory loops
  if (remove_self_edges) {
    edge_df <- edge_df %>% filter(gene != perturbation)
  }

  # Keep valid genes and perturbations
  edge_df <- edge_df %>% filter(gene %in% genes, perturbation %in% perts)

  # Construct adjacency matrix
  edge_mat <- sparseMatrix(i = match(edge_df$gene, genes), 
                           j = match(edge_df$perturbation, perts),
                           x = rep(TRUE, nrow(edge_df)),
                           dims = c(length(genes), length(perts)),
                           dimnames = list(genes, perts))

  # In-degree
  target_in_degree <- as.integer(Matrix::rowSums(edge_mat != 0))

  # Out-degree
  perturbation_out_degree <- as.integer(Matrix::colSums(edge_mat != 0))

  # In-degree distribution
  in_degree_table <- tibble(degree = target_in_degree) %>%
    count(degree, name = "freq") %>%
    arrange(degree) %>%
    mutate(prob = freq / sum(freq))

  # Out-degree distribution
  out_degree_table <- tibble(degree = perturbation_out_degree) %>%
    count(degree, name = "freq") %>%
    arrange(degree) %>%
    mutate(prob = freq / sum(freq))

  list(
    posterior_cutoff = posterior_cutoff,
    edge_df = edge_df,
    edge_mat = edge_mat,
    target_in_degree = target_in_degree,
    perturbation_out_degree = perturbation_out_degree,
    in_degree_table = in_degree_table,
    out_degree_table = out_degree_table
  )
}

posterior_grid <- c(0.5, 0.9, 0.95)

degree_results <- setNames(
  lapply(
    posterior_grid,
    make_edges_and_degrees,
    all_de = all_de,
    genes = genes,
    perts = perts,
    remove_self_edges = FALSE
  ),
  paste0("Posterior_", posterior_grid)
)

# Degree Distributions
degree_dist_all <- bind_rows(lapply(names(degree_results), function(nm) {

  in_tab <- degree_results[[nm]]$in_degree_table %>%
    mutate(Posterior = nm, degree_type = "In-degree: target gene j")

  out_tab <- degree_results[[nm]]$out_degree_table %>%
    mutate(Posterior = nm, degree_type = "Out-degree: perturbation gene k")

  bind_rows(in_tab, out_tab)
}))

# Plotting
in_degree_raw_all <- bind_rows(lapply(names(degree_results), function(nm) {
  tibble(degree = degree_results[[nm]]$target_in_degree, Posterior = nm)
}))

ggplot(in_degree_raw_all, aes(x = degree)) +
  geom_histogram(binwidth = 1, boundary = -0.5) +
  facet_wrap(~ Posterior, scales = "free_y") +
  labs(
    title = "Norman in-degree distribution",
    subtitle = "Edges defined by limma empirical Bayes analysis with BH-adjusted p-values",
    x = "Degree of target gene j",
    y = "Frequency"
  ) +
  theme_classic()

out_degree_raw_all <- bind_rows(lapply(names(degree_results), function(nm) {
  tibble(degree = degree_results[[nm]]$perturbation_out_degree, Posterior = nm)
}))

ggplot(out_degree_raw_all, aes(x = degree)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ Posterior, scales = "free_y") +
  labs(
    title = "Norman out-degree distribution",
    subtitle = "Edges defined by limma empirical Bayes analysis with BH-adjusted p-values",
    x = "Degree of perturbation gene k",
    y = "Frequency"
  ) +
  theme_classic()
```

For each FDR cutoff $\tau \in\{0.01, 0.05, 0.15\}$, define the binary perturbation-target edge indicator:

$$
\delta_{kj} = \mathbf 1\{q_{kj}\le \tau\}
$$\
Where:

-   $k = 1,\dots,K$ indexes perturbation regulators.

-   $j = 1,\dots,J$ indexes target genes.

-   $q_{kj}$ is the BH-adjusted p-value from the limma empirical Bayes analysis.

-   $\delta_{kj}=1$ means perturbation $k$ significantly affects target gene $j$, and $\delta_{kj}=0$ otherwise.

Define target in-degree and perturbation out-degree as

$$
c_j=\sum_{k=1}^{K}\delta_{kj}, \qquad j=1,\dots,J,
$$

$$
d_k=\sum_{j=1}^{J}\delta_{kj}, \qquad k=1,\dots,K.
$$

Define in-degree vector for target $j$ and out-degree vector for perturbation $k$ as

$$
\boldsymbol{\delta}_{·j} = (\delta_{1j}, \delta_{2j}, \ldots, \delta_{Kj}), \qquad \boldsymbol{\delta}_{k·} = (\delta_{k1}, \delta_{k2}, \ldots, \delta_{kJ}).
$$

Define full perturbation-target edge matrix as

$$
\mathbf D=(\delta_{kj})_{K\times J}.
$$

## **1. Bayesian Selection Framework**

For each candidate model $M$, the posterior model probability is calculated using Bayes' theorem:

$$
p(M\mid \mathbf D) = \frac{p(\mathbf D\mid M)p(M)}{p(\mathbf D)}
$$

where

$$
p(\mathbf D)={\sum_{M}p(\mathbf D\mid M)p(M)}.
$$

Here, $\mathbf D$ is the data used by model $M$. The marginal likelihood for model $M$ is

$$
p(\mathbf D\mid M) = \int_\theta p(\mathbf D\mid \theta,M)p(\theta\mid M)d\theta.
$$

After obtaining the model evidence estimate for each candidate model, we combine all model evidence estimates using Bayes' theorem.

## **2. Marginal Models**

The candidate models are:

-   $M_\text{BB}$: Beta-Binomial model

-   Geometric model

    -   $M_\text{GU}$: Untruncated geometric model

    -   $M_\text{GT}$: Truncated geometric model

-   $M_\text{SZ}$: Shifted Zipf model

### **2.1** $M_\text{BB}$: **Beta-Binomial Model**

#### **2.1.1 Target degree** $c_j$

$$
c_j \mid \phi_j, M_{\text{BB,in}} \sim \operatorname{Binomial}(K, \phi_j), \quad \phi_j \mid M_{\text{BB,in}}\sim \text{Beta}(\alpha,\beta).
$$

**Model evidence:**

Equivalently,

$$
\delta_{kj}\mid \phi_j, M_{\text{BB,in}}\sim \operatorname{Bernoulli}(\phi_j),
$$

so

$$
p(\delta_{kj}\mid \phi_j, M_{\text{BB,in}}) = \phi_j^{\delta_{kj}} (1-\phi_j)^{1-\delta_{kj}}.
$$

Define $\boldsymbol{\phi} = (\phi_1,\ldots,\phi_J)$ and condition on $\boldsymbol{\phi}$, the likelihood of the edge matrix $\mathbf D=(\delta_{kj})_{K\times J}$ is

$$
p(\mathbf D\mid\boldsymbol{\phi},M_{\text{BB,in}}) = \prod_{j=1}^{J} \prod_{k=1}^{K} \phi_j^{\delta_{kj}} (1-\phi_j)^{1-\delta_{kj}}
$$

$$
= \prod_{j=1}^{J} \phi_j^{c_j} (1-\phi_j)^{K-c_j}.
$$

Integrating out $\boldsymbol{\phi}$,

$$
p(\mathbf D \mid M_{\text{BB,in}}) = \int_{[0,1]^J} p(\mathbf{D} \mid \boldsymbol{\phi},M_{\text{BB,in}}) p(\boldsymbol{\phi} \mid M_{\text{BB,in}}) \, d\boldsymbol{\phi}
$$

$$
= \prod_{j=1}^{J} \frac{1}{B(\alpha,\beta)} \int_0^1 \phi_j^{c_j} (1-\phi_j)^{K-c_j} \phi_j^{\alpha-1} (1-\phi_j)^{\beta-1}  \,d\phi_j
$$

$$
= \prod_{j=1}^{J} \frac{1}{B(\alpha,\beta)} \int_0^1 \phi_j^{\alpha+c_j-1} (1-\phi_j)^{\beta+K-c_j-1} \,d\phi_j
$$

$$
= \prod_{j=1}^{J} \frac{ B(\alpha+c_j,\beta+K-c_j) }{ B(\alpha,\beta) } .
$$

**Posterior predictive:**

Using Bayes' theorem,

$$
p(\boldsymbol{\phi}\mid\mathbf D,M_{\text{BB,in}}) = \frac{p(\mathbf D\mid\boldsymbol{\phi },M_{\text{BB,in}})p(\boldsymbol{\phi }\mid M_{\text{BB,in}})} {p(\mathbf D \mid M_{\text{BB,in}})}
$$

$$
= \frac{\prod_{j=1}^{J} \frac{1}{B(\alpha,\beta)} \phi_j^{\alpha+c_j-1} (1-\phi_j)^{\beta+K-c_j-1}}{\prod_{j=1}^{J} \frac{B(\alpha+c_j,\beta+K-c_j)}{B(\alpha,\beta)}}
$$

$$
= \prod_{j=1}^{J} \frac{\phi_j^{\alpha+c_j-1} (1-\phi_j)^{\beta+K-c_j-1}}{B(\alpha+c_j,\beta+K-c_j)}
$$

$$
 = \prod_{j=1}^{J} \operatorname{Beta} \left(\phi_j \mid \alpha+c_j, \beta+K-c_j \right).
$$

For a new edge $\tilde{\delta}_{kj}$,

$$
p(\tilde{\delta}_{kj}\mid \mathbf D,M_{\text{BB,in}}) = \int p(\tilde{\delta}_{kj}\mid\boldsymbol{\phi},M_{\text{BB,in}}) p(\boldsymbol{\phi}\mid\mathbf D,M_{\text{BB,in}}) \,d\boldsymbol{\phi}
$$

$$
= \frac{1} {B(\alpha+c_j,\beta+K-c_j)} \int_0^1 \phi_j^{\tilde{\delta}_{kj}} (1-\phi_j)^{1-\tilde{\delta}_{kj}} \phi_j^{\alpha+c_j-1} (1-\phi_j)^{\beta+K-c_j-1} \,d\phi_j
$$

$$
= \frac{1}{B(\alpha+c_j, \beta+K-c_j)} \int_0^1 \phi_j^{\alpha+c_j+\tilde{\delta}_{kj}-1} (1-\phi_j)^{\beta+K-c_j+1-\tilde{\delta}_{kj}-1} \, d\phi_j
$$

$$
= \frac{ B( \alpha+c_j+\tilde{\delta}_{kj}, \beta+K-c_j+1-\tilde{\delta}_{kj} ) }{ B(\alpha+c_j,\beta+K-c_j) }.
$$

When $\tilde{\delta}_{kj}=1$,

$$
p(\tilde{\delta}_{kj}=1\mid\mathbf D,M_{\text{BB,in}}) = \frac{ B(\alpha+c_j+1,\beta+K-c_j) }{ B(\alpha+c_j,\beta+K-c_j) }
$$

$$
 = \frac{\alpha+c_j} {\alpha+\beta+K} .
$$

#### **2.1.2 Perturbation degree** $d_k$

$$
d_k\mid \psi_k, M_{\text{BB,out}}\sim\operatorname{Binomial}(J,\psi_k), \qquad \psi_k \mid M_{\text{BB,out}}\sim\operatorname{Beta}(\alpha,\beta).
$$

**Model evidence:**

Equivalently,

$$
\delta_{kj}\mid\psi_k,M_{\text{BB,out}} \sim \operatorname{Bernoulli}(\psi_k).
$$

Following the same derivation as for the in-degree model,

$$
p(\mathbf D \mid M_{\text{BB,out}}) = \prod_{k=1}^{K} \frac{ B(\alpha+d_k,\beta+J-d_k) }{ B(\alpha,\beta) } .
$$

**Posterior predictive:**

For a new edge $\tilde{\delta}_{kj}$,

$$
p(\tilde{\delta}_{kj}=1\mid\mathbf D, M_{\text{BB,out}}) = \frac{\alpha+d_k} {\alpha+\beta+J} .
$$

### **2.2** **Geometric Model**

#### **2.2.1** $M_\text{GU}$: **Untruncated Geometric**

**2.2.1.1 Target degree** $c_j$

$$
c_j \mid \lambda, M_{\text{GU,in}}\sim\operatorname{Geometric}(\lambda), \qquad \lambda \mid M_{\text{GU,in}}\sim\operatorname{Beta}(\alpha,\beta),
$$

**Model evidence:**

$$
p(c_j\mid\lambda,M_{\text{GU,in}}) = (1-\lambda)^{c_j}\lambda, \qquad c_j=0,1,2,\ldots.
$$

We define the incoming edge vector for target gene $j$ as $\boldsymbol{\delta}_{\cdot j} = (\delta_{1j},\ldots,\delta_{Kj})$. Assume that the edge configurations with the same in-degree $c_j$ are exchangeable. Since there are $\binom{K}{c_j}$ possible edge configurations with exactly $c_j$ edges,

$$
p(\boldsymbol{\delta}_{\cdot j}\mid c_j,M_{\text{GU,in}}) = {\binom{K}{c_j}}^{-1}.
$$

Since $c_j$ is a deterministic function of $\boldsymbol{\delta}_{\cdot j}$,

$$
p(\boldsymbol{\delta}_{\cdot j}\mid\lambda,M_{\text{GU,in}}) = p(\boldsymbol{\delta}_{\cdot j},c_j\mid\lambda,M_{\text{GU,in}})
$$

$$
= p(\boldsymbol{\delta}_{\cdot j}\mid c_j,M_{\text{GU,in}}) p(c_j\mid\lambda,M_{\text{GU,in}})
$$

$$
= \binom{K}{c_j}^{-1} (1-\lambda)^{c_j}\lambda.
$$

Conditional on $\lambda$, the likelihood of the edge matrix $\mathbf D=(\delta_{kj})_{K\times J}$ is

$$p(\mathbf D\mid\lambda,M_{\text{GU,in}}) = \prod_{j=1}^{J} p(\boldsymbol{\delta}_{\cdot j}\mid\lambda,M_{\text{GU,in}})$$

$$
= \prod_{j=1}^{J} \binom{K}{c_j}^{-1} (1-\lambda)^{c_j}\lambda.
$$

Let $S = \sum_{k=1}^{K}\sum_{j=1}^{J}\delta_{kj} = \sum_{j=1}^{J}c_j$ represent the total number of edges, then

$$
p(\mathbf D\mid\lambda,M_{\text{GU,in}}) = \left[ \prod_{j=1}^{J} \binom{K}{c_j} \right]^{-1}(1-\lambda)^S \lambda^J .
$$

Integrating out $\lambda$,

$$
p(\mathbf D \mid M_{\text{GU,in}}) = \int_0^1 p(\mathbf D\mid\lambda,M_{\text{GU,in}})p(\lambda \mid M_{\text{GU,in}}) \,d\lambda
$$

$$
= \int_0^1  \left[ \prod_{j=1}^{J} \binom{K}{c_j} \right]^{-1} (1-\lambda)^S \lambda^J \frac{\lambda^{\alpha-1} (1-\lambda)^{\beta-1}}{B(\alpha,\beta)}  \,d\lambda
$$

$$
= \left[ \prod_{j=1}^{J} \binom{K}{c_j} \right]^{-1} \frac{1}{B(\alpha,\beta)} \int_0^1  (1-\lambda)^S \lambda^J\lambda^{\alpha-1} (1-\lambda)^{\beta-1} \,d\lambda
$$

$$
= \left[ \prod_{j=1}^{J} \binom{K}{c_j} \right]^{-1} \frac{1}{B(\alpha,\beta)} \int_0^1 \lambda^{\alpha+J-1} (1-\lambda)^{\beta+S-1} \,d\lambda
$$

$$
= \left[ \prod_{j=1}^{J} \binom{K}{c_j} \right]^{-1} \frac{ B(\alpha+J,\beta+S) }{ B(\alpha,\beta) } .
$$

**Posterior predictive:**

Using Bayes' theorem,

$$
p(\lambda\mid\mathbf D,M_{\text{GU,in}}) = \frac{ p(\mathbf D\mid\lambda,M_{\text{GU,in}})p(\lambda \mid M_{\text{GU,in}}) }{ p(\mathbf D \mid M_{\text{GU,in}}) }
$$

$$
=\frac{\left[ \prod_{j=1}^{J} \binom{K}{c_j} \right]^{-1}(1-\lambda)^S \lambda^J \frac{\lambda^{\alpha-1} (1-\lambda)^{\beta-1}}{B(\alpha,\beta)}}{\left[ \prod_{j=1}^{J} \binom{K}{c_j} \right]^{-1} \frac{ B(\alpha+J,\beta+S) }{ B(\alpha,\beta) }}
$$

$$
= \frac{ \lambda^{\alpha+J-1} (1-\lambda)^{\beta+S-1} }{ B(\alpha+J,\beta+S) }
$$

$$
=\operatorname{Beta} (\alpha+J,\beta+S) .
$$

For a new edge $\tilde{\delta}_{kj}$, since all edge configurations with the same $\tilde c_j$ are exchangeable,

$$
p(\tilde{\delta}_{kj}=1\mid\tilde c_j,M_{\text{GU,in}}) = \frac{\tilde c_j}{K}.
$$

Therefore, since $\tilde{c}_j$ is discrete,

$$
p(\tilde{\delta}_{kj}=1\mid\lambda,M_{\text{GU,in}}) = \sum_{\tilde{c}_j=0}^{\infty} p(\tilde{\delta}_{kj}=1\mid\tilde{c}_j,M_{\text{GU,in}}) p(\tilde{c}_j\mid\lambda,M_{\text{GU,in}})
$$

$$
= \frac{1}{K} E[\tilde c_j\mid\lambda]
$$

$$
= \frac{1}{K} \frac{1-\lambda}{\lambda}.
$$

Integrating over the posterior distribution of $\lambda$,

$$
p(\tilde{\delta}_{kj}=1\mid\mathbf D,M_{\text{GU,in}}) = \int_0^1 p(\tilde{\delta}_{kj}=1\mid\lambda,M_{\text{GU,in}}) p(\lambda\mid\mathbf D,M_{\text{GU,in}}) \,d\lambda
$$

$$
= \frac{1}{K} \frac{1} {B(\alpha+J,\beta+S)} \int_0^1 \frac{1-\lambda}{\lambda} \lambda^{\alpha+J-1} (1-\lambda)^{\beta+S-1} \,d\lambda
$$

$$
= \frac{1}{K} \frac{1} {B(\alpha+J,\beta+S)} \int_0^1 \lambda^{\alpha+J-2} (1-\lambda)^{\beta+S} \,d\lambda
$$

$$
= \frac{1}{K} \frac{ B(\alpha+J-1,\beta+S+1) }{ B(\alpha+J,\beta+S) }
$$

$$
= \frac{\beta+S} {K(\alpha+J-1)}.
$$

**2.2.1.2 Perturbation degree** $d_k$

$$
d_k \mid  \lambda, M_{\text{GU,out}}\sim\operatorname{Geometric}(\lambda), \qquad \lambda \mid M_{\text{GU,out}}\sim\operatorname{Beta}(\alpha,\beta).
$$

**Model evidence:**

Following the same derivation as for the target in-degree model,

$$
p(\mathbf D\mid\lambda,M_{\text{GU,out}}) = \left[ \prod_{k=1}^{K} \binom{J}{d_k} \right]^{-1} (1-\lambda)^S \lambda^K.
$$

Integrating out $\lambda$,

$$
p(\mathbf D \mid M_{\text{GU,out}}) = \left[ \prod_{k=1}^{K} \binom{J}{d_k} \right]^{-1} \frac{ B(\alpha+K,\beta+S) }{ B(\alpha,\beta) } .
$$

**Posterior predictive:**

$$
p(\lambda\mid\mathbf D,M_{\text{GU,out}}) =\operatorname{Beta} (\alpha+K,\beta+S)
$$

For a new edge $\tilde{\delta}_{kj}$,

$$
p(\tilde{\delta}_{kj}=1\mid\lambda,M_{\text{GU,out}}) = \frac{1}{J} E[\tilde d_k\mid\lambda] = \frac{1}{J} \frac{1-\lambda}{\lambda}.
$$

Integrating over the posterior distribution of $\lambda$,

$$
p(\tilde{\delta}_{kj}=1\mid\mathbf D,M_{\text{GU,out}}) = \frac{\beta+S} {J(\alpha+K-1)} .
$$

#### **2.2.2** $M_\text{GT}$: **Truncated Geometric**

**2.2.2.1 Target degree** $c_j$

Since $c_j\in\{0,1,\dots,K\}$, we use a truncated geometric distribution on $\{0,\dots,K\}$. The ordinary zero-supported geometric distribution is

$$
p(c_j=k\mid\lambda,M_{\text{GT,in}}) = \lambda(1-\lambda)^k, \qquad k=0,1,2,\dots,
$$

$$
\lambda \mid M_{\text{GT,in}} \sim \operatorname{Beta}(\alpha, \beta).
$$

However, in this problem the degree cannot exceed $K$. Therefore, we condition the ordinary geometric distribution on $c_j\le K$:

$$
p(c_j=k \mid c_j \le K, \lambda,M_{\text{GT,in}}) = \frac{\lambda(1-\lambda)^k}{\sum_{\ell=0}^{K} \lambda(1-\lambda)^\ell}
$$

$$
= \frac{(1-\lambda)^k}{\sum_{\ell=0}^{K}(1-\lambda)^\ell}.
$$

Thus,

$$
p(c_j=k\mid\lambda, M_{\text{GT,in}}) = \frac{(1-\lambda)^k}{\sum_{\ell=0}^{K} (1-\lambda)^\ell} \qquad k=0,\dots,K,
$$

As in the untruncated geometric model, assume that all edge configurations with the same in-degree $c_j$ are exchangeable. Then,

$$
p(\boldsymbol{\delta}_{\cdot j}\mid c_j,M_{\text{GT,in}}) = \binom{K}{c_j}^{-1}.
$$

Because $c_j$ is a deterministic function of $\boldsymbol{\delta}_{\cdot j}$,

$$
p(\boldsymbol{\delta}_{\cdot j}\mid\lambda,M_{\text{GT,in}}) = p(\boldsymbol{\delta}_{\cdot j}\mid c_j,M_{\text{GT,in}}) p(c_j\mid\lambda,M_{\text{GT,in}})
$$

$$
= \binom{K}{c_j}^{-1} \frac{(1-\lambda)^{c_j}}{\sum_{\ell=0}^{K}(1-\lambda)^\ell}.
$$

Assuming target genes are conditionally independent given $\lambda$, the likelihood of the edge matrix is

$$
p(\mathbf D\mid\lambda,M_{\text{GT,in}}) = \prod_{j=1}^{J} p(\boldsymbol{\delta}_{\cdot j}\mid\lambda,M_{\text{GT,in}})
$$

$$
= \prod_{j=1}^{J} \left[ \binom{K}{c_j}^{-1} \frac{(1-\lambda)^{c_j}}{\sum_{\ell=0}^{K}(1-\lambda)^\ell} \right].
$$

The model evidence is therefore

$$
p(\mathbf D\mid M_{\text{GT,in}}) = \int_0^1 p(\mathbf D\mid\lambda,M_{\text{GT,in}}) p(\lambda\mid M_{\text{GT,in}}) \,d\lambda
$$

$$
= \int_0^1 \prod_{j=1}^{J} \left[ \binom{K}{c_j}^{-1} \frac{(1-\lambda)^{c_j}} {\sum_{\ell=0}^{K}(1-\lambda)^\ell} \right] \frac{ \lambda^{\alpha-1}(1-\lambda)^{\beta-1} }{ B(\alpha,\beta) } \,d\lambda .
$$

This integral does not reduce to the conjugate closed form and is therefore evaluated numerically using Stan and bridge sampling.

**2.2.2.2 Perturbation degree** $d_k$

Since $d_k\in\{0,1,\dots,J\}$, the truncated geometric model is

$$
p(d_k=r\mid\lambda,M_{\text{GT,out}}) = \frac{(1-\lambda)^r}{\sum_{\ell=0}^{J}(1-\lambda)^\ell}, \qquad r=0,\dots,J,
$$

$$
\lambda\mid M_{\mathrm{GT,out}} \sim \operatorname{Beta}(\alpha,\beta). 
$$

Assume that all edge configurations with the same out-degree $d_k$ are exchangeable, then

$$
p(\boldsymbol{\delta}_{k\cdot}\mid d_k,M_{\text{GT,out}}) = \binom{J}{d_k}^{-1}.
$$

Therefore,

$$
p(\boldsymbol{\delta}_{k\cdot}\mid\lambda,M_{\text{GT,out}}) = \binom{J}{d_k}^{-1} \frac{(1-\lambda)^{d_k}}{\sum_{\ell=0}^{J}(1-\lambda)^\ell}.
$$

Assuming perturbations are conditionally independent given $\lambda$,

$$
p(\mathbf D\mid\lambda,M_{\text{GT,out}}) = \prod_{k=1}^{K} \left[ \binom{J}{d_k}^{-1} \frac{(1-\lambda)^{d_k}}{\sum_{\ell=0}^{J}(1-\lambda)^\ell} \right].
$$

The model evidence is

$$
p(\mathbf D\mid M_{\text{GT,out}}) = \int_0^1 p(\mathbf D\mid\lambda,M_{\text{GT,out}}) p(\lambda\mid M_{\text{GT,out}}) \,d\lambda
$$

$$
= \int_0^1 \prod_{k=1}^{K} \left[ \binom{J}{d_k}^{-1} \frac{(1-\lambda)^{d_k}} {\sum_{\ell=0}^{J}(1-\lambda)^\ell} \right] \frac{ \lambda^{\alpha-1}(1-\lambda)^{\beta-1} }{ B(\alpha,\beta) } \,d\lambda. 
$$

### **2.3** $M_\text{SZ}$: **Shifted Zipf Model**

#### **2.3.1 Target degree** $c_j$

Since $c_j$ can be zero, Zipf's law is applied to $c_j+1$. The truncated shifted Zipf distribution is

$$
p(c_j=k\mid s,M_{\text{SZ,in}}) = \frac{(k+1)^{-s}}{\sum_{\ell=0}^{K}(\ell+1)^{-s}}, \qquad k=0,\dots,K.
$$

As above, assume that all edge configurations with the same in-degree $c_j$ are exchangeable:

$$
p(\boldsymbol{\delta}_{\cdot j}\mid c_j,M_{\text{SZ,in}}) = \binom{K}{c_j}^{-1}.
$$

Therefore,

$$
p(\boldsymbol{\delta}_{\cdot j}\mid s,M_{\text{SZ,in}}) = p(\boldsymbol{\delta}_{\cdot j}\mid c_j,M_{\text{SZ,in}}) p(c_j\mid s,M_{\text{SZ,in}})
$$

$$
= \binom{K}{c_j}^{-1} \frac{(c_j+1)^{-s}}{\sum_{\ell=0}^{K}(\ell+1)^{-s}}.
$$

Assuming target genes are conditionally independent given $s$,

$$
p(\mathbf D\mid s,M_{\text{SZ,in}}) = \prod_{j=1}^{J} \left[ \binom{K}{c_j}^{-1} \frac{(c_j+1)^{-s}}{\sum_{\ell=0}^{K}(\ell+1)^{-s}} \right].
$$

The model evidence is

$$p(\mathbf D\mid M_{\text{SZ,in}}) = \int p(\mathbf D\mid s,M_{\text{SZ,in}}) p(s\mid M_{\text{SZ,in}}) \,ds$$

$$
= \int \prod_{j=1}^{J} \left[ \binom{K}{c_j}^{-1} \frac{(c_j+1)^{-s}}{\sum_{\ell=0}^{K}(\ell+1)^{-s}} \right] p(s\mid M_{\text{SZ,in}}) \,ds.
$$

This evidence is evaluated using Stan and bridge sampling.

#### **2.3.2 Perturbation degree** $d_k$

For the perturbation out-degree,

$$
p(d_k=r\mid s,M_{\text{SZ,out}}) = \frac{(r+1)^{-s}}{\sum_{\ell=0}^{J}(\ell+1)^{-s}}, \qquad r=0,\dots,J.
$$

Assume that all outgoing edge configurations with the same out-degree $d_k$ are exchangeable:

$$
p(\boldsymbol{\delta}_{k\cdot}\mid d_k,M_{\text{SZ,out}}) = \binom{J}{d_k}^{-1}.
$$

Therefore,

$$
p(\boldsymbol{\delta}_{k\cdot}\mid s,M_{\text{SZ,out}}) = \binom{J}{d_k}^{-1} \frac{(d_k+1)^{-s}}{\sum_{\ell=0}^{J}(\ell+1)^{-s}}.
$$

Assuming perturbations are conditionally independent given $s$,

$$
p(\mathbf D\mid s,M_{\text{SZ,out}}) = \prod_{k=1}^{K} \left[ \binom{J}{d_k}^{-1} \frac{(d_k+1)^{-s}}{\sum_{\ell=0}^{J}(\ell+1)^{-s}} \right].
$$

The model evidence is

$$
p(\mathbf D\mid M_{\text{SZ,out}}) = \int p(\mathbf D\mid s,M_{\text{SZ,out}}) p(s\mid M_{\text{SZ,out}}) \,ds
$$

$$
= \int \prod_{k=1}^{K} \left[ \binom{J}{d_k}^{-1} \frac{(d_k+1)^{-s}}{\sum_{\ell=0}^{J}(\ell+1)^{-s}} \right] p(s\mid M_{\text{SZ,out}}) \,ds.
$$

## **3. Bernoulli Models**

The three Bernoulli models are:

-   $M_\text{HBI}$: Independent Heterogeneous Bernoulli Model

-   $M_\text{HBG}$: Global Heterogeneous Bernoulli Model

-   $M_\text{HBRF1}$: Rank-1 Factorized Heterogeneous Bernoulli Model

### **3.1** $M_\text{HBI}$: **Independent Heterogeneous Bernoulli Model**

$$
\delta_{kj} \mid \pi_{kj},M_{\text{HBI}}
\overset{\text{ind}}{\sim}
\operatorname{Bernoulli}(\pi_{kj}),
\qquad
\pi_{kj}\mid M_{\text{HBI}}\sim\operatorname{Beta}(\alpha,\beta).
$$

**Model evidence:**

Integrating out edge-specific probabilities $\pi_{kj}$,

$$
p(\delta_{kj} \mid M_{\text{HBI}})=
\int
p(\delta_{kj}\mid\pi_{kj},M_{\text{HBI}})
p(\pi_{kj} \mid M_{\text{HBI}})
\,d\pi_{kj}
$$

$$
=\int
\pi_{kj}^{\delta_{kj}}
(1-\pi_{kj})^{1-\delta_{kj}}
\frac {\pi_{kj}^{\alpha-1}
(1-\pi_{kj})^{\beta-1}}{B(\alpha,\beta)}
\,d\pi_{kj}
$$

$$
= \frac{1}{B(\alpha,\beta)}
\int
\pi_{kj}^{\alpha+\delta_{kj}-1}
(1-\pi_{kj})^{\beta+1-\delta_{kj}-1}
\,d\pi_{kj}
$$

$$
= \frac{B(\alpha+\delta_{kj},\beta+1-\delta_{kj})}{B(\alpha,\beta)}.
$$

When $\delta_{kj}=1$,

$$
p(\delta_{kj}=1\mid M_{\text{HBI}})=
\frac{B(\alpha+1,\beta)}{B(\alpha,\beta)} = \frac{\alpha}{\alpha+\beta}.
$$

When $\delta_{kj}=0$,

$$
p(\delta_{kj}=0 \mid M_{\text{HBI}})=
\frac{\beta}{\alpha+\beta}.
$$

Because different $\pi_{kj}$'s are independent, so marginally,

$$
\delta_{kj} \mid M_{\text{HBI}}
\overset{\text{ind}}{\sim}
\operatorname{Bernoulli}
\left(
\frac{\alpha}{\alpha+\beta}
\right).
$$

Let $S=\sum_k\sum_j\delta_{kj}$ represent the total number of edges, then

$$
p(\mathbf D\mid M_{\text{HBI}})=
\prod_{k=1}^{K} \prod_{j=1}^{J}p(\delta_{kj} \mid M_{\text{HBI}})=
\left(
\frac{\alpha}{\alpha+\beta}
\right)^S
\left(
\frac{\beta}{\alpha+\beta}
\right)^{KJ-S}.
$$

**Posterior predictive:**

For a newly generated edge $\tilde{\delta}_{kj}$ with a newly drawn independent edge-specific probability $\tilde{\pi}_{kj}$, the observed matrix $\mathbf D$ provides no information about $\tilde{\pi}_{kj}$. Therefore, the prior predictive and posterior predictive distributions are identical:

$$
p(\tilde \delta_{kj}​ = 1∣\mathbf D,M_{\text{HBI}})=p(\tilde \delta_{kj}​ = 1 \mid M_{\text{HBI}})​ = \frac{\alpha}{\alpha+\beta}.
$$

### **3.2** $M_\text{HBG}$: **Global Heterogeneous Bernoulli Model**

$$
\delta_{kj} \mid \pi, M_{\text{HBG}}\sim\operatorname{Bernoulli}(\pi), \quad \pi\mid M_{\text{HBG}}\sim\operatorname{Beta}(\alpha,\beta).
$$

**Model evidence:**

$$
p(\mathbf D\mid M_{\text{HBG}}) = \int_0^1  p(\mathbf D\mid\pi,M_{\text{HBG}})p(\pi\mid M_{\text{HBG}})\,d\pi
$$

$$
= \int_0^1  \left[ \prod_{k=1}^{K}\prod_{j=1}^{J} \pi^{\delta_{kj}} (1-\pi)^{1-\delta_{kj}} \right] \frac{\pi^{\alpha-1}(1-\pi)^{\beta-1}} {B(\alpha,\beta)} \,d\pi
$$

$$
= \frac{1}{B(\alpha,\beta)} \int_0^1 \pi^S(1-\pi)^{KJ-S} \pi^{\alpha-1}(1-\pi)^{\beta-1} \,d\pi
$$

$$
= \frac{1}{B(\alpha,\beta)} \int_0^1 \pi^{\alpha+S-1} (1-\pi)^{\beta+KJ-S-1} \,d\pi
$$

$$
= \frac{B(\alpha+S,\beta+KJ-S)} {B(\alpha,\beta)}.
$$

**Posterior predictive:**

Using Bayes' theorem,

$$
p(\pi\mid\mathbf D,M_{\text{HBG}}) = \frac{p(\mathbf D\mid\pi,M_{\text{HBG}})p(\pi \mid M_{\text{HBG}})} {p(\mathbf D \mid M_{\text{HBG}})}
$$

$$
= \frac{B(\alpha, \beta)}{B(\alpha+S, \beta+KJ-S)} \frac{1} {B(\alpha, \beta)} \pi^{\alpha+S-1} (1-\pi)^{\beta+KJ-S-1}
$$

$$
= \frac{1}{B(\alpha+S, \beta+KJ-S)} \pi^{\alpha+S-1} (1-\pi)^{\beta+KJ-S-1}
$$

$$
=\operatorname{Beta} (\alpha+S,\beta+KJ-S). 
$$

Thus, for a new edge $\tilde{\delta}_{kj}$,

$$
p(\tilde{\delta}_{kj}\mid\mathbf D, M_{\text{HBG}}) = \int_0^1  p(\tilde{\delta}_{kj}\mid\pi, M_{\text{HBG}}) p(\pi\mid\mathbf D, M_{\text{HBG}}) \,d\pi
$$

$$
= \frac{1}{B(\alpha+S,\beta+KJ-S)} \int_0^1 \pi^{\tilde{\delta}_{kj}} (1-\pi)^{1-\tilde{\delta}_{kj}} \pi^{\alpha+S-1} (1-\pi)^{\beta+KJ-S-1} \,d\pi
$$

$$
= \frac{1}{B(\alpha+S,\beta+KJ-S)} \int_0^1 \pi^{\alpha+S+\tilde{\delta}_{kj}-1} (1-\pi)^{\beta+KJ-S+1-\tilde{\delta}_{kj}-1} \,d\pi
$$

$$
= \frac{ B(\alpha+S+\tilde{\delta}_{kj}, \beta+KJ-S+1-\tilde{\delta}_{kj}) }{ B(\alpha+S,\beta+KJ-S) }.
$$

When $\tilde{\delta}_{kj}=1$,

$$
p(\tilde{\delta}_{kj}=1\mid\mathbf D, M_{\text{HBG}}) = \frac{ B(\alpha+S+1,\beta+KJ-S) }{ B(\alpha+S,\beta+KJ-S) }
$$

$$
=\frac{\alpha+S} {\alpha+\beta+KJ}.
$$

### **3.3** $M_\text{HBRF1}$: **Rank-1 Factorized Heterogeneous Bernoulli Model**

$$
\delta_{kj} \mid \pi_{kj}, M_{\text{HBRF1}}\overset{\text{ind}}{\sim} \operatorname{Bernoulli}(\pi_{kj}), \qquad k=1,\ldots,K,\quad j=1,\ldots,J
$$

$$
\pi_{kj}=\psi_k\phi_j, \quad \psi_k \mid M_{\text{HBRF1}}\sim \operatorname{Beta}(\alpha_\psi,\beta_\psi), \quad \phi_j \mid M_{\text{HBRF1}}\sim \operatorname{Beta}(\alpha_\phi,\beta_\phi).
$$

Here, $\psi_k$ represents the perturbation-specific contribution to the edge probability, whereas $\phi_j$ represents the target-specific contribution.

Define the perturbation-factor vector and target-factor vector as

$$
 \boldsymbol\psi=(\psi_1,\ldots,\psi_K), \quad \boldsymbol\phi=(\phi_1,\ldots,\phi_J).
$$

Conditional on $\boldsymbol{\psi}$ and $\boldsymbol{\phi}$, the edges $\delta_{kj}$ are independent. However, because edges in the same row share $\psi_k$ and edges in the same column share $\phi_j$, they are not marginally independent after the latent variables are integrated out.

#### **3.3.1 Target degree** $c_j$

To obtain a simpler marginal representation of the in-degree distribution, we condition on $\phi_j$ and integrate out $\psi_k$:

$$
p(\delta_{kj}=1\mid\phi_j, M_{\text{HBRF1}}) = \int_0^1 p(\delta_{kj}=1\mid\psi_k,\phi_j, M_{\text{HBRF1}}) p(\psi_k \mid M_{\text{HBRF1}}) \,d\psi_k
$$

$$
= \int_0^1 \psi_k\phi_j p(\psi_k \mid M_{\text{HBRF1}}) \,d\psi_k
$$

$$
= E[\psi_k]\phi_j.
$$

Since

$$
E[\psi_k] = \int_0^1 \psi_k \cdot \left[ \frac{1}{B(\alpha_\psi, \beta_\psi)} \, \psi_k^{\alpha_\psi - 1} (1 - \psi_k)^{\beta_\psi - 1} \right] d\psi_k
$$

$$
= \frac{B(\alpha_\psi + 1, \beta_\psi)}{B(\alpha_\psi, \beta_\psi)}
$$

$$
= \frac{\alpha_\psi} {\alpha_\psi+\beta_\psi},
$$

we define $\mu_\psi = E[\psi_k] = \frac{\alpha_\psi} {\alpha_\psi+\beta_\psi}$, then

$$
p(\delta_{kj}=1\mid\phi_j, M_{\text{HBRF1}}) = \mu_\psi\phi_j.
$$

Therefore, the in-degree model can be represented by the simplified model

$$
c_j\mid\phi_j, M_{\text{HBRF1S,in}} \sim \operatorname{Binomial} (K,\mu_\psi\phi_j), 
$$

$$
\qquad \mu_\psi = \frac{\alpha_\psi} {\alpha_\psi+\beta_\psi}, \qquad \phi_j \mid M_{\text{HBRF1S,in}}\sim \operatorname{Beta}(\alpha_\phi,\beta_\phi).
$$

Hence,

$$
p(c_j\mid\phi_j, M_{\text{HBRF1S,in}}) = \binom{K}{c_j} (\mu_\psi\phi_j)^{c_j} (1-\mu_\psi\phi_j)^{K-c_j}.
$$

Conditional on $c_j$, all $\binom{K}{c_j}$ edge configurations are exchangeable. Therefore

$$
p(\boldsymbol{\delta}_{\cdot j} \mid c_j,M_{\text{HBRF1S,in}}) = \binom{K}{c_j}^{-1}.
$$

Because $c_j$ is a deterministic function of $\boldsymbol{\delta}_{\cdot j}$,

$$
p(\boldsymbol{\delta}_{\cdot j} \mid \phi_j, M_{\text{HBRF1S,in}}) = p(\boldsymbol{\delta}_{\cdot j} \mid c_j, M_{\text{HBRF1S,in}}) p(c_j \mid \phi_j, M_{\text{HBRF1S,in}}) 
$$

$$
= (\mu_{\psi} \phi_j)^{c_j} (1 - \mu_{\psi} \phi_j)^{K - c_j}.
$$

Conditional on $\boldsymbol{\phi}$, the likelihood of the edge matrix $\mathbf D=(\delta_{kj})_{K\times J}$ is

$$
p(\mathbf{D} \mid \boldsymbol{\phi}, M_{\mathrm{HBRF1S,in}}) = \prod_{j=1}^{J} (\mu_\psi \phi_j)^{c_j} (1 - \mu_\psi \phi_j)^{K - c_j}
$$

Integrating out $\boldsymbol{\phi}$,

$$
p(\mathbf{D} \mid M_{\mathrm{HBRF1S,in}}) = \int_{[0,1]^J} p(\mathbf{D} \mid \boldsymbol{\phi}, M_{\mathrm{HBRF1S,in}}) \, p(\boldsymbol{\phi} \mid M_{\mathrm{HBRF1S,in}}) \, d\boldsymbol{\phi}
$$

$$
= \int_{[0,1]^J} \prod_{j=1}^{J} \left[ (\mu_\psi \phi_j)^{c_j} (1 - \mu_\psi \phi_j)^{K - c_j} \frac{\phi_j^{\alpha_\phi - 1} (1 - \phi_j)^{\beta_\phi - 1}}{B(\alpha_\phi, \beta_\phi)} \right] d\boldsymbol{\phi}
$$

$$
=\prod_{j=1}^{J} \int_0^1 (\mu_\psi \phi_j)^{c_j} (1 - \mu_\psi \phi_j)^{K - c_j} \frac{\phi_j^{\alpha_\phi - 1} (1 - \phi_j)^{\beta_\phi - 1}}{B(\alpha_\phi, \beta_\phi)} \, d\phi_j.
$$

This integral does not reduce to the conjugate closed form and is therefore evaluated numerically using Stan and bridge sampling.

#### **3.3.2 Perturbation degree** $d_k$

Similarly, we define $\mu_\phi = E[\phi_j] = \frac{\alpha_\phi} {\alpha_\phi+\beta_\phi}$ and condition on $\psi_k$,

$$
p(\delta_{kj}=1\mid\psi_k, M_{\text{HBRF1}}) = \mu_\phi\psi_k.
$$

Therefore, the out-degree model can also be represented by the simplified model

$$
d_k\mid\psi_k,M_{\text{HBRF1S,out}} \sim \operatorname{Binomial} (J,\mu_\phi\psi_k), 
$$

$$
\mu_\phi = \frac{\alpha_\phi} {\alpha_\phi+\beta_\phi}, \quad \psi_k \mid M_{\text{HBRF1S,out}}\sim \operatorname{Beta}(\alpha_\psi,\beta_\psi).
$$

Exchangeability across the $\binom{J}{d_k}$ row configurations gives:

$$
p(\boldsymbol{\delta}_{k \cdot} \mid \psi_k, M_{\mathrm{HBRF1S,out}}) = (\mu_\phi \psi_k)^{d_k} (1 - \mu_\phi \psi_k)^{J - d_k}
$$

Conditioned on $\boldsymbol{\psi} = (\psi_1, \ldots, \psi_K)$, row edge vectors $\boldsymbol{\delta}_{1 \cdot}, \ldots, \boldsymbol{\delta}_{K \cdot}$ are independent:

$$
p(\mathbf{D} \mid \boldsymbol{\psi}, M_{\mathrm{HBRF1S,out}}) = \prod_{k=1}^{K} (\mu_\phi \psi_k)^{d_k} (1 - \mu_\phi \psi_k)^{J - d_k}
$$

Integrating out $\boldsymbol{\psi}$,

$$
p(\mathbf{D} \mid M_{\mathrm{HBRF1S,out}}) = \prod_{k=1}^{K} \int_0^1 (\mu_\phi \psi_k)^{d_k} (1 - \mu_\phi \psi_k)^{J - d_k} \frac{\psi_k^{\alpha_\psi - 1} (1 - \psi_k)^{\beta_\psi - 1}}{B(\alpha_\psi, \beta_\psi)} \, d\psi_k.
$$

## **4. Model Fit**

Model fit is assessed using two complementary criteria:

-   **Model evidence:** for models with analytically tractable marginal likelihoods, the derived closed-form expressions are used directly without Stan. For models without closed-form solutions, bridge sampling is applied to the fitted Stan model to estimate the marginal likelihood.

-   **Posterior predictive checks:** posterior predictive distributions are obtained either directly from the analytical posterior distributions or from posterior draws generated by Stan.

The in-degree and out-degree marginal models are fitted separately. In particular, the in-degree and out-degree likelihoods are not multiplied because the degree vectors $\mathbf c=(c_1,\ldots,c_J)$ and $\mathbf d=(d_1,\ldots,d_K)$ are both determined by the same edge matrix $\mathbf D$ and are therefore dependent.

*Note:* For hyperparameter selection of $M_\text{GT}$ and $M_\text{HBRF1S}$, the marginal likelihood is evaluated by direct numerical integration over the specified hyperparameter grid. After selecting the evidence-maximizing hyperparameters, Stan is used for posterior inference and bridge sampling is used for the final marginal-likelihood estimate.

```{r}
fdr_name <- "FDR_0.01"
fdr_label <- sub("FDR_", "", fdr_name)

# D = K perturbations x J target genes
D <- t(as.matrix(degree_results[[fdr_name]]$edge_mat))
storage.mode(D) <- "integer"

K <- nrow(D)
J <- ncol(D)

# Target in-degree c_j
c_obs <- as.integer(colSums(D))

# Perturbation out-degree d_k
d_obs <- as.integer(rowSums(D))

# Total number of edges
S_obs <- sum(D)
```

```{r}
# Closed-form model evidence
# M_BB,in
log_evidence_BB_in <- function(alpha, beta, c, K) {
  sum(lbeta(alpha + c, beta + K - c) - lbeta(alpha, beta))
}

# M_BB,out
log_evidence_BB_out <- function(alpha, beta, d, J) {
  sum(lbeta(alpha + d, beta + J - d) - lbeta(alpha, beta))
}

# M_GU,in
log_evidence_GU_in <- function(alpha, beta, c, K) {
  
  J <- length(c)
  S <- sum(c)

  -sum(lchoose(K, c)) + lbeta(alpha + J, beta + S) - lbeta(alpha, beta)
}

# M_GU,out
log_evidence_GU_out <- function(alpha, beta, d, J) {

  K <- length(d)
  S <- sum(d)

  -sum(lchoose(J, d)) + lbeta(alpha + K, beta + S) - lbeta(alpha, beta)
}

# M_HBG
log_evidence_HBG <- function(alpha, beta, S, K, J) {
  lbeta(alpha + S, beta + K * J - S) - lbeta(alpha, beta)
}

# Optimize the closed-form Beta hyperparameters: mu in [0.1, 0.9], kappa in [1, 100]
optimize_hyperparameters <- function(log_evidence, ...) {

  opt <- optim(par = c(mu = 0.5, kappa = 10),

    fn = function(par) {
      mu <- par[1]
      kappa <- par[2]

      alpha <- mu * kappa
      beta <- (1 - mu) * kappa

      -do.call(log_evidence, c(list(alpha = alpha, beta = beta), list(...)))
    },

    method = "L-BFGS-B",
    lower = c(0.1, 1),
    upper = c(0.9, 100)
  )

  mu <- opt$par[1]
  kappa <- opt$par[2]

  tibble(mu = mu, kappa = kappa, alpha = mu * kappa, 
         beta = (1 - mu) * kappa, log_evidence = -opt$value)
}

# M_BB,in
opt_BB_in <- optimize_hyperparameters(log_evidence = log_evidence_BB_in, 
                                      c = c_obs, K = K)

# M_BB,out
opt_BB_out <- optimize_hyperparameters(log_evidence = log_evidence_BB_out, 
                                       d = d_obs, J = J)

# M_GU,in
opt_GU_in <- optimize_hyperparameters(log_evidence = log_evidence_GU_in, 
                                      c = c_obs, K = K)

# M_GU,out
opt_GU_out <- optimize_hyperparameters(log_evidence = log_evidence_GU_out, 
                                       d = d_obs, J = J)

# M_HBG
opt_HBG <- optimize_hyperparameters(log_evidence = log_evidence_HBG, 
                                    S = S_obs, K = K, J = J)

# M_HBI
opt_HBI_raw <- optimize(f = function(mu) {
  -(S_obs * log(mu) + (K * J - S_obs) * log(1 - mu))
  }, 
  interval = c(0.1, 0.9))
opt_HBI <- tibble(mu = opt_HBI_raw$minimum, log_evidence = -opt_HBI_raw$objective)
```

```{r}
# Truncated Geometric
dir.create("stan_models", showWarnings = FALSE)

writeLines('
data {
  int<lower=1> K;
  int<lower=1> J;
  array[K, J] int<lower=0, upper=1> D;

  // 1 = in-degree, 2 = out-degree
  int<lower=1, upper=2> degree_type;

  real<lower=0> alpha;
  real<lower=0> beta;
}

transformed data {
  array[J] int c;
  array[K] int d;
  int S;

  // c_j = sum_k delta_kj
  for (j in 1:J) {
    c[j] = 0;

    for (k in 1:K) {
      c[j] += D[k, j];
    }
  }

  // d_k = sum_j delta_kj
  for (k in 1:K) {
    d[k] = 0;

    for (j in 1:J) {
      d[k] += D[k, j];
    }
  }

  // S = total number of edges
  S = 0;

  for (j in 1:J) {
    S += c[j];
  }
}

parameters {
  real<lower=0, upper=1> lambda;
}

model {
  real log_one_minus_lambda;

  target += beta_lpdf(lambda | alpha, beta);

  log_one_minus_lambda = log1m(lambda);

  // M_GT,in
  if (degree_type == 1) {

    vector[K + 1] log_terms;
    real log_config;

    log_config = 0;

    for (j in 1:J) {
      log_config += lchoose(K, c[j]);
    }

    for (l in 0:K) {
      log_terms[l + 1] = l * log_one_minus_lambda;
    }

    target += -log_config + S * log_one_minus_lambda - J * log_sum_exp(log_terms);
  }

  // M_GT,out
  if (degree_type == 2) {

    vector[J + 1] log_terms;
    real log_config;

    log_config = 0;

    for (k in 1:K) {
      log_config += lchoose(J, d[k]);
    }

    for (l in 0:J) {
      log_terms[l + 1] = l * log_one_minus_lambda;
    }

    target += -log_config + S * log_one_minus_lambda - K * log_sum_exp(log_terms);
  }
}
', "stan_models/truncated_geometric.stan")

stan_GT <- stan_model("stan_models/truncated_geometric.stan")

# log model evidence for the hyperparameter grid search
log_evidence_GT <- function(alpha, beta, degree, max_degree) {

  N <- length(degree)
  S <- sum(degree)

  # Probability of the specific observed edge configuration
  log_config <- -sum(lchoose(max_degree, degree))

  # Log integrand after lambda = logistic(t)
  logf <- function(t) {

    lambda <- plogis(t)
    log_lambda <- plogis(t, log.p = TRUE)
    log1m <- plogis(t, lower.tail = FALSE, log.p = TRUE)

    # Z(lambda)
    logZ <- numeric(length(t))

    at_zero <- lambda == 0
    at_one <- lambda == 1
    middle <- !(at_zero | at_one)

    # lim lambda -> 0: Z = max_degree + 1
    logZ[at_zero] <- log(max_degree + 1)

    # lim lambda -> 1: Z = 1
    logZ[at_one] <- 0

    logZ[middle] <- log(-expm1((max_degree + 1) * log1m[middle])) - log_lambda[middle]
    alpha * log_lambda + (beta + S) * log1m - N * logZ - lbeta(alpha, beta)
  }

  # Scaling constant only for numerical stability
  mode <- optimize(
    function(t) -logf(t),
    interval = c(-100, 100)
  )
  m <- -mode$objective

  val <- integrate(
    function(t) {
      exp(logf(t) - m)
    },
    lower = -Inf,
    upper = Inf,
    subdivisions = 2000,
    rel.tol = 1e-8
  )$value

  log_config + m + log(val)
}

# Hyperparameter grid search because model evidence is not closed form
mu_grid <- seq(0.1, 0.9, by = 0.1)
kappa_grid <- c(1, 3, 10, 30, 100)

hp_grid <- tidyr::expand_grid(mu = mu_grid, kappa = kappa_grid) %>%
  mutate(alpha = mu * kappa, beta = (1 - mu) * kappa)

# M_GT,in
GT_in_grid <- hp_grid %>%
  rowwise() %>%
  mutate(log_evidence =log_evidence_GT(alpha = alpha, beta = beta,
                                       degree = c_obs, max_degree = K)) %>%
  ungroup() %>%
  arrange(desc(log_evidence))

best_GT_in <- GT_in_grid %>%
  slice(1)

# M_GT,out
GT_out_grid <- hp_grid %>%
  rowwise() %>%
  mutate(log_evidence = log_evidence_GT(alpha = alpha, beta = beta,
                                        degree = d_obs, max_degree = J)) %>%
  ungroup() %>%
  arrange(desc(log_evidence))

best_GT_out <- GT_out_grid %>%
  slice(1)

# Fit M_GT using optimized hyperparameters
stan_GT_in_data <- list(K = K, J = J, D = D, degree_type = 1L, 
                        alpha = best_GT_in$alpha[[1]], beta = best_GT_in$beta[[1]])
fit_GT_in <- sampling(stan_GT, data = stan_GT_in_data, iter = 10000, warmup = 2000, 
                      chains = 3, seed = 123, cores = 1)
bridge_GT_in <- bridge_sampler(fit_GT_in, silent = TRUE)
logml_GT_in <- bridge_GT_in$logml

stan_GT_out_data <- list(K = K, J = J, D = D, degree_type = 2L, 
                         alpha = best_GT_out$alpha[[1]], beta = best_GT_out$beta[[1]])
fit_GT_out <- sampling(stan_GT, data = stan_GT_out_data, iter = 10000, warmup = 2000,
                       chains = 3, seed = 123, cores = 1)
bridge_GT_out <- bridge_sampler(fit_GT_out, silent = TRUE)
logml_GT_out <- bridge_GT_out$logml

# MCMC convergence
summary(fit_GT_in)$summary[, c("n_eff", "Rhat")]
summary(fit_GT_out)$summary[, c("n_eff", "Rhat")]
```

```{r}
# Shifted Zipf
writeLines('
data {
  int<lower=1> K;
  int<lower=1> J;
  array[K, J] int<lower=0, upper=1> D;

  // 1 = in-degree, 2 = out-degree
  int<lower=1, upper=2> degree_type;
}

transformed data {
  array[J] int c;
  array[K] int d;

  // c_j = sum_k delta_kj
  for (j in 1:J) {
    c[j] = 0;

    for (k in 1:K) {
      c[j] += D[k, j];
    }
  }

  // d_k = sum_j delta_kj
  for (k in 1:K) {
    d[k] = 0;

    for (j in 1:J) {
      d[k] += D[k, j];
    }
  }
}

parameters {
  real<lower=1e-6, upper=10> s;
}

model {
  target += uniform_lpdf(s | 1e-6, 10);

  // M_SZ,in
  if (degree_type == 1) {

    vector[K + 1] log_terms;
    real log_config;
    real sum_log;

    log_config = 0;
    sum_log = 0;

    for (j in 1:J) {
      log_config += lchoose(K, c[j]);
      sum_log += log(c[j] + 1);
    }

    for (l in 0:K) {
      log_terms[l + 1] = -s * log(l + 1);
    }

    target += -log_config -s * sum_log -J * log_sum_exp(log_terms);
  }

  // M_SZ,out
  if (degree_type == 2) {

    vector[J + 1] log_terms;
    real log_config;
    real sum_log;

    log_config = 0;
    sum_log = 0;

    for (k in 1:K) {
      log_config += lchoose(J, d[k]);
      sum_log += log(d[k] + 1);
    }

    for (l in 0:J) {
      log_terms[l + 1] = -s * log(l + 1);
    }

    target += -log_config -s * sum_log -K * log_sum_exp(log_terms);
  }
}
', "stan_models/shifted_zipf.stan")

stan_SZ <- stan_model("stan_models/shifted_zipf.stan")

# M_SZ,in
stan_SZ_in_data <- list(K = K, J = J, D = D, degree_type = 1L)
fit_SZ_in <- sampling(stan_SZ, data = stan_SZ_in_data, iter = 10000, warmup = 2000, chains = 3, seed = 123, cores = 1)
bridge_SZ_in <- bridge_sampler(fit_SZ_in, silent = TRUE)
logml_SZ_in <- bridge_SZ_in$logml

# M_SZ,out
stan_SZ_out_data <- list(K = K, J = J, D = D, degree_type = 2L)
fit_SZ_out <- sampling(stan_SZ, data = stan_SZ_out_data, iter = 10000, warmup = 2000, chains = 3, seed = 123, cores = 1)
bridge_SZ_out <- bridge_sampler(fit_SZ_out, silent = TRUE)
logml_SZ_out <- bridge_SZ_out$logml

# MCMC convergence
summary(fit_SZ_in)$summary[, c("n_eff", "Rhat")]
summary(fit_SZ_out)$summary[, c("n_eff", "Rhat")]
```

```{r}
# HBRF1S in-degree
writeLines('
data {
  int<lower=1> K;
  int<lower=1> J;

  array[K, J] int<lower=0, upper=1> D;

  // mu_psi = E[psi_k]
  real<lower=0, upper=1> mu_psi;

  // phi_j ~ Beta(alpha_phi, beta_phi)
  real<lower=1e-9> alpha_phi;
  real<lower=1e-9> beta_phi;
}

transformed data {
  array[J] int c;

  // c_j = sum_k delta_kj
  for (j in 1:J) {

    c[j] = 0;

    for (k in 1:K) {
      c[j] += D[k, j];
    }
  }
}

parameters {
  vector<lower=0, upper=1>[J] phi;
}

model {

  // phi_j ~ Beta(alpha_phi, beta_phi)
  target += beta_lpdf(phi | alpha_phi, beta_phi);

  for (j in 1:J) {

    real p_j;

    p_j = mu_psi * phi[j];

    target += binomial_lpmf(c[j] | K, p_j) - lchoose(K, c[j]);
  }
}
', "stan_models/HBRF1S_in.stan")

stan_HBRF1S_in <- stan_model("stan_models/HBRF1S_in.stan")

# HBRF1S log model evidence for hyperparameter grid search
# In-degree: mu_factor = mu_psi, alpha = alpha_phi, beta = beta_phi, degree = c_obs, max_degree = K
# Out-degree: mu_factor = mu_phi, alpha = alpha_psi, beta = beta_psi, degree = d_obs, max_degree = J
log_evidence_HBRF1S <- function(mu_factor, alpha, beta, degree, max_degree) {

  degree_table <- table(degree)
  degree_value <- as.integer(names(degree_table))
  degree_freq <- as.numeric(degree_table)

  # Marginal log probability for one specific edge configuration
  log_marginal_one_degree <- function(x) {

    logf <- function(t) {

      theta <- plogis(t)

      # Numerically stable log(theta) and log(1-theta)
      log_theta <- plogis(t, log.p = TRUE)
      log_one_minus_theta <- plogis(t, lower.tail = FALSE, log.p = TRUE)

      # Exact log integrand after theta = logistic(t)
      (
        x * log(mu_factor) +
          (x + alpha) * log_theta +
          beta * log_one_minus_theta +
          (max_degree - x) *
            log1p(-mu_factor * theta) -
          lbeta(alpha, beta)
      )
    }

    # Find maximum only for numerical scaling
    mode <- optimize(
      function(t) -logf(t),
      interval = c(-100, 100)
    )
    m <- -mode$objective

    # Integrate
    val <- integrate(
      function(t) {
        exp(logf(t) - m)
      },
      lower = -Inf,
      upper = Inf,
      subdivisions = 2000,
      rel.tol = 1e-8
    )$value

    # Restore
    m + log(val)
  }

  log_prob <- vapply(degree_value, log_marginal_one_degree, numeric(1))
  sum(degree_freq * log_prob)
}

# Hyperparameter grid for HBRF1S in-degree
# alpha_phi = mu_phi * kappa_phi
# beta_phi  = (1 - mu_phi) * kappa_phi
phi_hp_grid <- tidyr::expand_grid(mu_phi = mu_grid, kappa_phi = kappa_grid) %>%
  mutate(alpha_phi = mu_phi * kappa_phi, beta_phi = (1 - mu_phi) * kappa_phi)
HBRF1S_in_hp_grid <- tidyr::crossing(mu_psi = mu_grid, phi_hp_grid)

# M_HBRF1S,in hyperparameter grid search
HBRF1S_in_grid <- HBRF1S_in_hp_grid %>%
  rowwise() %>%
  mutate(log_evidence = log_evidence_HBRF1S(mu_factor = mu_psi, alpha = alpha_phi,
                                            beta = beta_phi, degree = c_obs, 
                                            max_degree = K)) %>%
  ungroup() %>%
  arrange(desc(log_evidence))

# Select hyperparameters with highest model evidence
best_HBRF1S_in <- HBRF1S_in_grid %>%
  slice(1)

# Fit HBRF1S in-degree using optimized hyperparameters
stan_HBRF1S_in_data <- list(K = K, J = J, D = D,
                            mu_psi = best_HBRF1S_in$mu_psi[[1]],
                            alpha_phi = best_HBRF1S_in$alpha_phi[[1]],
                            beta_phi = best_HBRF1S_in$beta_phi[[1]])

fit_HBRF1S_in <- sampling(stan_HBRF1S_in, data = stan_HBRF1S_in_data, iter = 10000,
                          warmup = 2000, chains = 3, seed = 123, cores = 1)

# Save to run bridge_sampler on server
# saveRDS(fit_HBRF1S_in, "fit_HBRF1S_in.rds")
# saveRDS(stan_HBRF1S_in_data, "stan_HBRF1S_in_data.rds")

# bridge_HBRF1S_in <- bridge_sampler(fit_HBRF1S_in, silent = TRUE)
# logml_HBRF1S_in <- bridge_HBRF1S_in$logml

# MCMC convergence
# summary(fit_HBRF1S_in)$summary[, c("n_eff", "Rhat")]

# HBRF1S out-degree
writeLines('
data {
  int<lower=1> K;
  int<lower=1> J;

  array[K, J] int<lower=0, upper=1> D;

  // mu_phi = E[phi_j]
  real<lower=0, upper=1> mu_phi;

  // psi_k ~ Beta(alpha_psi, beta_psi)
  real<lower=1e-9> alpha_psi;
  real<lower=1e-9> beta_psi;
}

transformed data {
  array[K] int d;

  // d_k = sum_j delta_kj
  for (k in 1:K) {

    d[k] = 0;

    for (j in 1:J) {
      d[k] += D[k, j];
    }
  }
}

parameters {
  vector<lower=0, upper=1>[K] psi;
}

model {

  // psi_k ~ Beta(alpha_psi, beta_psi)
  target += beta_lpdf(psi | alpha_psi, beta_psi);

  for (k in 1:K) {

    real p_k;

    p_k = mu_phi * psi[k];

    target += binomial_lpmf(d[k] | J, p_k) - lchoose(J, d[k]);
  }
}
', "stan_models/HBRF1S_out.stan")

stan_HBRF1S_out <- stan_model("stan_models/HBRF1S_out.stan")

# Hyperparameter grid for HBRF1S out-degree
# alpha_psi = mu_psi * kappa_psi
# beta_psi  = (1 - mu_psi) * kappa_psi
psi_hp_grid <- tidyr::expand_grid(mu_psi = mu_grid, kappa_psi = kappa_grid) %>%
  mutate(alpha_psi = mu_psi * kappa_psi, beta_psi = (1 - mu_psi) * kappa_psi)
HBRF1S_out_hp_grid <- tidyr::crossing(mu_phi = mu_grid, psi_hp_grid)

# M_HBRF1S,out hyperparameter grid search
HBRF1S_out_grid <- HBRF1S_out_hp_grid %>%
  rowwise() %>%
  mutate(log_evidence = log_evidence_HBRF1S(mu_factor = mu_phi, 
                                            alpha = alpha_psi, beta = beta_psi,
                                            degree = d_obs, max_degree = J)) %>%
  ungroup() %>%
  arrange(desc(log_evidence))

# Select hyperparameters with highest model evidence
best_HBRF1S_out <- HBRF1S_out_grid %>%
  slice(1)

# Fit HBRF1S out-degree using optimized hyperparameters
stan_HBRF1S_out_data <- list(K = K, J = J, D = D,
                             mu_phi = best_HBRF1S_out$mu_phi[[1]],
                             alpha_psi = best_HBRF1S_out$alpha_psi[[1]],
                             beta_psi = best_HBRF1S_out$beta_psi[[1]])

fit_HBRF1S_out <- sampling(stan_HBRF1S_out, data = stan_HBRF1S_out_data,
                           iter = 10000, warmup = 2000, chains = 3, seed = 123, cores = 1)

bridge_HBRF1S_out <- bridge_sampler(fit_HBRF1S_out, silent = TRUE)
logml_HBRF1S_out <- bridge_HBRF1S_out$logml

# MCMC convergence
summary(fit_HBRF1S_out)$summary[, c("n_eff", "Rhat")]
```

```{r}
# Model comparison
# log p(D | M)
logml <- c(
  M_BB_in = opt_BB_in$log_evidence[[1]],
  M_GU_in = opt_GU_in$log_evidence[[1]],
  M_GT_in = bridge_GT_in$logml,
  M_SZ_in = bridge_SZ_in$logml,

  M_BB_out = opt_BB_out$log_evidence[[1]],
  M_GU_out = opt_GU_out$log_evidence[[1]],
  M_GT_out = bridge_GT_out$logml,
  M_SZ_out = bridge_SZ_out$logml,

  M_HBI = opt_HBI$log_evidence[[1]],
  M_HBG = opt_HBG$log_evidence[[1]],

  M_HBRF1S_in = best_HBRF1S_in$log_evidence[[1]],
  M_HBRF1S_out = bridge_HBRF1S_out$logml
)

# log p(M) with same prior probability
log_prior <- log(rep(1 / length(logml), length(logml)))
names(log_prior) <- names(logml)

# log (p(D ∣ M)p(M))
log_bayes_numerator <- logml + log_prior

# log p(D)
log_bayes_denominator <- max(log_bayes_numerator) + log(sum(exp(log_bayes_numerator - max(log_bayes_numerator))))

# p(M | D)
posterior_model_probability <- exp(log_bayes_numerator - log_bayes_denominator)

# Comparison table
model_comparison <- tibble(

  model = names(logml),

  data_used = c(
    rep("D via c_j", 4),
    rep("D via d_k", 4),
    rep("D", 2),
    "D via c_j",
    "D via d_k"
  ),

  model_family = c(
    "M_BB,in: Beta-Binomial",
    "M_GU,in: Untruncated geometric",
    "M_GT,in: Truncated geometric",
    "M_SZ,in: Shifted Zipf",

    "M_BB,out: Beta-Binomial",
    "M_GU,out: Untruncated geometric",
    "M_GT,out: Truncated geometric",
    "M_SZ,out: Shifted Zipf",

    "M_HBI: Independent heterogeneous Bernoulli",
    "M_HBG: Global heterogeneous Bernoulli",

    "M_HBRF1S,in: Rank-1 factorized heterogeneous Bernoulli",
    "M_HBRF1S,out: Rank-1 factorized heterogeneous Bernoulli"
  ),

  log_marginal_likelihood = as.numeric(logml),
  prior_model_probability = exp(log_prior),
  posterior_model_probability = as.numeric(posterior_model_probability)) %>%
  arrange(desc(posterior_model_probability))

model_comparison
```

```{r}
# Posterior predictive checks 
set.seed(123)
ppc_degree_sample <- 1000

# Plotting function
plot_ppc_degree <- function(obs_degree, sim_degree, model_name, degree_type, 
                            max_degree, bins = 30) {

  # Stack posterior predictive degrees across PPC samples
  stacked_sim_degree <- as.vector(sim_degree)

  # Perturbation out-degree d_k
  if (degree_type == "Perturbation out-degree d_k") {

    # Common bins over the complete possible degree range
    plot_max_degree <- max(max_degree, obs_degree, stacked_sim_degree, na.rm = TRUE)
    breaks <- seq(-0.5, plot_max_degree + 0.5, length.out = bins + 1)

    # Observed histogram
    obs_hist <- hist(obs_degree, breaks = breaks, plot = FALSE, include.lowest = TRUE)

    # Posterior predictive histogram
    sim_hist <- hist(stacked_sim_degree, breaks = breaks, plot = FALSE, include.lowest = TRUE)

    # Convert counts to probabilities
    plot_df <- bind_rows(
      tibble(bin_mid = obs_hist$mids, prob = obs_hist$counts / sum(obs_hist$counts),
             source = "Observed"),
      tibble(bin_mid = sim_hist$mids, prob = sim_hist$counts / sum(sim_hist$counts),
             source = "Posterior predictive"))

    ggplot(plot_df, aes(x = bin_mid, y = prob, fill = source)) +
      geom_col(position = "identity", alpha = 0.45, width = diff(breaks)[1]) +
      labs(title = paste0(model_name, ", FDR = ", fdr_label),
           x = "Degree", y = "Probability", fill = NULL) +
      theme_classic()

  } else {

    # Target in-degree c_j
    plot_df <- bind_rows(
      tibble(degree = obs_degree, source = "Observed"),
      tibble(degree = stacked_sim_degree, source = "Posterior predictive")) %>%
      count(source, degree, name = "freq") %>%
      group_by(source) %>%
      mutate(prob = freq / sum(freq)) %>%
      ungroup()
    
    ggplot(plot_df, aes(x = degree, y = prob, fill = source)) +
      geom_col(position = "identity", alpha = 0.45, width = 1) +
      labs(title = paste0(model_name, ", FDR = ", fdr_label),
           x = "Degree", y = "Probability", fill = NULL) +
      theme_classic()
  }
}

# M_BB,in: analytical posterior
alpha_BB_in <- opt_BB_in$alpha[[1]]
beta_BB_in  <- opt_BB_in$beta[[1]]

sim_BB_infit_c <- matrix(NA, nrow = ppc_degree_sample, ncol = J)
sim_BB_infit_d <- matrix(NA, nrow = ppc_degree_sample, ncol = K)

for (s in seq_len(ppc_degree_sample)) {

  # Posterior phi_j from the in-degree fit
  phi_s <- rbeta(J, shape1 = alpha_BB_in + c_obs, shape2 = beta_BB_in + K - c_obs)

  # Full replicated K x J edge matrix with each column j has Bernoulli probability phi_j
  D_s <- matrix(rbinom(K * J, size = 1, prob = rep(phi_s, each = K)), nrow = K, ncol = J)

  sim_BB_infit_c[s, ] <- colSums(D_s)
  sim_BB_infit_d[s, ] <- rowSums(D_s)
}

qdraw(plot_ppc_degree(obs_degree = c_obs, sim_degree = sim_BB_infit_c,
                      model_name = "M_BB,in(in)",
                      degree_type = "Target in-degree c_j", max_degree = K))

qdraw(plot_ppc_degree(obs_degree = d_obs, sim_degree = sim_BB_infit_d,
                      model_name = "M_BB,in(out)",
                      degree_type = "Perturbation out-degree d_k", max_degree = J))

# M_BB,out: analytical posterior
alpha_BB_out <- opt_BB_out$alpha[[1]]
beta_BB_out  <- opt_BB_out$beta[[1]]

sim_BB_outfit_c <- matrix(NA, nrow = ppc_degree_sample, ncol = J)
sim_BB_outfit_d <- matrix(NA, nrow = ppc_degree_sample, ncol = K)

for (s in seq_len(ppc_degree_sample)) {

  # Posterior psi_k from out-degree fit
  psi_s <- rbeta(K, shape1 = alpha_BB_out + d_obs, shape2 = beta_BB_out + J - d_obs)

  # Full replicated edge matrix with each row k has Bernoulli probability psi_k
  D_s <- matrix(rbinom(K * J, size = 1, prob = rep(psi_s, times = J)), nrow = K, ncol = J)

  sim_BB_outfit_c[s, ] <- colSums(D_s)
  sim_BB_outfit_d[s, ] <- rowSums(D_s)
}

qdraw(plot_ppc_degree(obs_degree = c_obs, sim_degree = sim_BB_outfit_c,
                      model_name = "M_BB,out(in)",
                      degree_type = "Target in-degree c_j", max_degree = K))

qdraw(plot_ppc_degree(obs_degree = d_obs, sim_degree = sim_BB_outfit_d,
                      model_name = "M_BB,out(out)",
                      degree_type = "Perturbation out-degree d_k", max_degree = J))

# M_GU,in: analytical posterior
alpha_GU_in <- opt_GU_in$alpha[[1]]
beta_GU_in  <- opt_GU_in$beta[[1]]

sim_GU_in <- matrix(NA, nrow = ppc_degree_sample, ncol = J)
sim_GU_infit_d <- matrix(NA, nrow = ppc_degree_sample, ncol = K)

# Track how often exceeding happens
GU_in_exceed <- 0

for (s in seq_len(ppc_degree_sample)) {

  # lambda | D
  lambda_s <- rbeta(1, shape1 = alpha_GU_in + J, shape2 = beta_GU_in + S_obs)

  # Untruncated geometric, resample degrees exceeding K
  c_rep <- rgeom(J, prob = lambda_s)
  
  while (any(c_rep > K)) {
    idx_bad <- which(c_rep > K)
    GU_in_exceed <- GU_in_exceed + length(idx_bad)
    c_rep[idx_bad] <- rgeom(length(idx_bad), prob = lambda_s)
  }

  sim_GU_in[s, ] <- c_rep

  D_s <- matrix(0L, K, J)
  for (j in seq_len(J)) {
    if (c_rep[j] > 0) {
      D_s[sample.int(K, c_rep[j]), j] <- 1L
    }
  }
  sim_GU_infit_d[s, ] <- rowSums(D_s)
  
}

GU_in_exceed

qdraw(plot_ppc_degree(obs_degree = c_obs, sim_degree = sim_GU_in, 
                      model_name = "M_GU,in(in)",
                      degree_type = "Target in-degree c_j", max_degree = K))

qdraw(plot_ppc_degree(obs_degree = d_obs, sim_degree = sim_GU_infit_d,
                      model_name = "M_GU,in(out)",
                      degree_type = "Perturbation out-degree d_k", max_degree = J))

# M_GU,out: analytical posterior
alpha_GU_out <- opt_GU_out$alpha[[1]]
beta_GU_out  <- opt_GU_out$beta[[1]]

sim_GU_out <- matrix(NA, nrow = ppc_degree_sample, ncol = K)
sim_GU_outfit_c <- matrix(NA, nrow = ppc_degree_sample, ncol = J)

# Track how often exceeding happens
GU_out_exceed <- 0

for (s in seq_len(ppc_degree_sample)) {

  # lambda | D
  lambda_s <- rbeta(1, shape1 = alpha_GU_out + K, shape2 = beta_GU_out + S_obs)

  # Untruncated geometric, resample degrees exceeding J
  d_rep <- rgeom(K, prob = lambda_s)
  
  while (any(d_rep > J)) {
    idx_bad <- which(d_rep > J)
    GU_out_exceed <- GU_out_exceed + length(idx_bad)
    d_rep[idx_bad] <- rgeom(length(idx_bad), prob = lambda_s)
  }

  sim_GU_out[s, ] <- d_rep

  D_s <- matrix(0L, K, J)
  for (k in seq_len(K)) {
    if (d_rep[k] > 0) {
      D_s[k, sample.int(J, d_rep[k])] <- 1L
    }
  }
  sim_GU_outfit_c[s, ] <- colSums(D_s)
}

GU_out_exceed

qdraw(plot_ppc_degree(obs_degree = c_obs, sim_degree = sim_GU_outfit_c,
                      model_name = "M_GU,out(in)",
                      degree_type = "Target in-degree c_j", max_degree = K))

qdraw(plot_ppc_degree(obs_degree = d_obs, sim_degree = sim_GU_out,
                      model_name = "M_GU,out(out)",
                      degree_type = "Perturbation out-degree d_k", max_degree = J))

# Construct D^rep from a vector of in-degrees
make_D_from_in_degree <- function(c_rep, K, J) {

  D_s <- matrix(0L, K, J)

  for (j in seq_len(J)) {
    
    if (c_rep[j] > 0) {
      
      D_s[sample.int(K, c_rep[j]), j] <- 1L
      
    }
  }

  D_s
}

# Construct D^rep from a vector of out-degrees
make_D_from_out_degree <- function(d_rep, K, J) {

  D_s <- matrix(0L, K, J)

  for (k in seq_len(K)) {

    if (d_rep[k] > 0) {

      D_s[k, sample.int(J, d_rep[k])] <- 1L
      
    }
  }

  D_s
}

# M_GT,in: stan posterior draws
lambda_GT_in <- rstan::extract(fit_GT_in, pars = "lambda", permuted = TRUE)$lambda
idx <- sample(seq_along(lambda_GT_in), ppc_degree_sample)

sim_GT_infit_c <- matrix(NA, nrow = ppc_degree_sample, ncol = J)
sim_GT_infit_d <- matrix(NA, nrow = ppc_degree_sample, ncol = K)

degree_grid <- 0:K

for (s in seq_len(ppc_degree_sample)) {

  lambda_s <- lambda_GT_in[idx[s]]

  # GT in-degree probabilities
  log_prob_s <- degree_grid * log1p(-lambda_s)
  log_prob_s <- log_prob_s - max(log_prob_s)
  prob_s <- exp(log_prob_s)
  prob_s <- prob_s / sum(prob_s)

  # Draw replicated in-degrees
  c_rep <- sample(degree_grid, size = J, replace = TRUE, prob = prob_s)

  # Convert those degrees into D^rep
  D_s <- make_D_from_in_degree(c_rep = c_rep, K = K, J = J)

  sim_GT_infit_c[s, ] <- colSums(D_s)
  sim_GT_infit_d[s, ] <- rowSums(D_s)
}

qdraw(plot_ppc_degree(obs_degree = c_obs, sim_degree = sim_GT_infit_c,
                      model_name = "M_GT,in(in)",
                      degree_type = "Target in-degree c_j", max_degree = K))

qdraw(plot_ppc_degree(obs_degree = d_obs, sim_degree = sim_GT_infit_d,
                      model_name = "M_GT,in(out)",
                      degree_type = "Perturbation out-degree d_k", max_degree = J))

# M_GT,out: stan posterior draws
lambda_GT_out <- rstan::extract(fit_GT_out, pars = "lambda", permuted = TRUE)$lambda
idx <- sample(seq_along(lambda_GT_out), ppc_degree_sample)

sim_GT_outfit_c <- matrix(NA, nrow = ppc_degree_sample, ncol = J)
sim_GT_outfit_d <- matrix(NA, nrow = ppc_degree_sample, ncol = K)

degree_grid <- 0:J

for (s in seq_len(ppc_degree_sample)) {

  lambda_s <- lambda_GT_out[idx[s]]

  # GT out-degree probabilities
  log_prob_s <- degree_grid * log1p(-lambda_s)
  log_prob_s <- log_prob_s - max(log_prob_s)
  prob_s <- exp(log_prob_s)
  prob_s <- prob_s / sum(prob_s)

  # Draw replicated out-degrees
  d_rep <- sample(degree_grid, size = K, replace = TRUE, prob = prob_s)

  # Convert those degrees into D^rep
  D_s <- make_D_from_out_degree(d_rep = d_rep, K = K, J = J)

  sim_GT_outfit_c[s, ] <- colSums(D_s)
  sim_GT_outfit_d[s, ] <- rowSums(D_s)
}

qdraw(plot_ppc_degree(obs_degree = c_obs, sim_degree = sim_GT_outfit_c,
                      model_name = "M_GT,out(in)",
                      degree_type = "Target in-degree c_j", max_degree = K))

qdraw(plot_ppc_degree(obs_degree = d_obs, sim_degree = sim_GT_outfit_d,
                      model_name = "M_GT,out(out)",
                      degree_type = "Perturbation out-degree d_k", max_degree = J))

# M_SZ,in: stan posterior draws
s_SZ_in <- rstan::extract(fit_SZ_in, pars = "s", permuted = TRUE)$s
idx <- sample(seq_along(s_SZ_in), ppc_degree_sample)

sim_SZ_infit_c <- matrix(NA, nrow = ppc_degree_sample, ncol = J)
sim_SZ_infit_d <- matrix(NA, nrow = ppc_degree_sample, ncol = K)

degree_grid <- 0:K

for (s in seq_len(ppc_degree_sample)) {

  s_par <- s_SZ_in[idx[s]]

  # SZ in-degree probabilities
  log_prob_s <- -s_par * log(degree_grid + 1)
  log_prob_s <- log_prob_s - max(log_prob_s)
  prob_s <- exp(log_prob_s)
  prob_s <- prob_s / sum(prob_s)

  # Draw replicated in-degrees
  c_rep <- sample(degree_grid, size = J, replace = TRUE, prob = prob_s)

  # Convert those degrees into D^rep
  D_s <- make_D_from_in_degree(c_rep, K, J)

  sim_SZ_infit_c[s, ] <- colSums(D_s)
  sim_SZ_infit_d[s, ] <- rowSums(D_s)
}

qdraw(plot_ppc_degree(obs_degree = c_obs, sim_degree = sim_SZ_infit_c,
                      model_name = "M_SZ,in(in)",
                      degree_type = "Target in-degree c_j", max_degree = K))

qdraw(plot_ppc_degree(obs_degree = d_obs, sim_degree = sim_SZ_infit_d,
                      model_name = "M_SZ,in(out)",
                      degree_type = "Perturbation out-degree d_k", max_degree = J))

# M_SZ,out: stan posterior draws
s_SZ_out <- rstan::extract(fit_SZ_out, pars = "s", permuted = TRUE)$s
idx <- sample(seq_along(s_SZ_out), ppc_degree_sample)

sim_SZ_outfit_c <- matrix(NA, nrow = ppc_degree_sample, ncol = J)
sim_SZ_outfit_d <- matrix(NA, nrow = ppc_degree_sample, ncol = K)

degree_grid <- 0:J

for (s in seq_len(ppc_degree_sample)) {

  s_par <- s_SZ_out[idx[s]]

  # SZ out-degree probabilities
  log_prob_s <- -s_par * log(degree_grid + 1)
  log_prob_s <- log_prob_s - max(log_prob_s)
  prob_s <- exp(log_prob_s)
  prob_s <- prob_s / sum(prob_s)

  # Draw replicated out-degrees
  d_rep <- sample(degree_grid, size = K, replace = TRUE, prob = prob_s)

  # Convert those degrees into D^rep
  D_s <- make_D_from_out_degree(d_rep, K, J)

  sim_SZ_outfit_c[s, ] <- colSums(D_s)
  sim_SZ_outfit_d[s, ] <- rowSums(D_s)
}

qdraw(plot_ppc_degree(obs_degree = c_obs, sim_degree = sim_SZ_outfit_c,
                      model_name = "M_SZ,out(in)",
                      degree_type = "Target in-degree c_j", max_degree = K))

qdraw(plot_ppc_degree(obs_degree = d_obs, sim_degree = sim_SZ_outfit_d,
                      model_name = "M_SZ,out(out)",
                      degree_type = "Perturbation out-degree d_k", max_degree = J))

# M_HBI: direct posterior predictive distribution
ppc_edge_sample <- 1000
mu_HBI <- opt_HBI$mu[[1]]

sim_HBI_c <- matrix(NA, nrow = ppc_edge_sample, ncol = J)
sim_HBI_d <- matrix(NA, nrow = ppc_edge_sample, ncol = K)

for (s in seq_len(ppc_edge_sample)) {

  # Replicated binary edge matrix
  D_s <- matrix(rbinom(K * J, size = 1, prob = mu_HBI), nrow = K, ncol = J)

  # Target in-degree
  sim_HBI_c[s, ] <- colSums(D_s)

  # Perturbation out-degree
  sim_HBI_d[s, ] <- rowSums(D_s)
}

qdraw(plot_ppc_degree(obs_degree = c_obs, sim_degree = sim_HBI_c,
                      model_name = "M_HBI(in)",
                      degree_type = "Target in-degree c_j", max_degree = K))

qdraw(plot_ppc_degree(obs_degree = d_obs, sim_degree = sim_HBI_d,
                      model_name = "M_HBI(out)",
                      degree_type = "Perturbation out-degree d_k", max_degree = J))

# M_HBG: analytical posterior
alpha_HBG <- opt_HBG$alpha[[1]]
beta_HBG  <- opt_HBG$beta[[1]]

sim_HBG_c <- matrix(NA, nrow = ppc_edge_sample, ncol = J)
sim_HBG_d <- matrix(NA, nrow = ppc_edge_sample, ncol = K)

for (s in seq_len(ppc_edge_sample)) {

  # Draw global pi from analytical posterior
  pi_s <- rbeta(1, shape1 = alpha_HBG + S_obs, shape2 = beta_HBG + K * J - S_obs)

  # Replicated edge matrix
  D_s <- matrix(rbinom(K * J, size = 1, prob = pi_s), nrow = K, ncol = J)

  # Target in-degree
  sim_HBG_c[s, ] <- colSums(D_s)

  # Perturbation out-degree
  sim_HBG_d[s, ] <- rowSums(D_s)
}

qdraw(plot_ppc_degree(obs_degree = c_obs, sim_degree = sim_HBG_c, 
                      model_name = "M_HBG(in)",
                      degree_type = "Target in-degree c_j", max_degree = K))

# Perturbation out-degree
qdraw(plot_ppc_degree(obs_degree = d_obs, sim_degree = sim_HBG_d,
                      model_name = "M_HBG(out)",
                      degree_type = "Perturbation out-degree d_k", max_degree = J))

# M_HBRF1S,in: stan posterior draws
mu_psi_HBRF1S <- best_HBRF1S_in$mu_psi[[1]]
phi_draws <- rstan::extract(fit_HBRF1S_in, pars = "phi", permuted = TRUE)$phi
idx <- sample(seq_len(nrow(phi_draws)), ppc_degree_sample)

sim_HBRF1S_in <- matrix(NA, nrow = length(idx), ncol = J)

for (s in seq_along(idx)) {

  # Entire posterior phi vector from one Stan iteration
  phi_s <- phi_draws[idx[s],]

  # Replicated target in-degrees
  sim_HBRF1S_in[s, ] <- rbinom(J, size = K, prob = mu_psi_HBRF1S * phi_s)
}

qdraw(plot_ppc_degree(obs_degree = c_obs, sim_degree = sim_HBRF1S_in,
                      model_name = "M_HBRF1S(in)",
                      degree_type = "Target in-degree c_j", max_degree = K))

# M_HBRF1S,out: stan posterior draws
mu_phi_HBRF1S <- best_HBRF1S_out$mu_phi[[1]]
psi_draws <- rstan::extract(fit_HBRF1S_out, pars = "psi", permuted = TRUE)$psi
idx <- sample(seq_len(nrow(psi_draws)), ppc_degree_sample)

sim_HBRF1S_out <- matrix(NA, nrow = length(idx), ncol = K)

for (s in seq_along(idx)) {

  # Entire posterior psi vector from one Stan iteration
  psi_s <- psi_draws[idx[s],]

  # Replicated perturbation out-degrees
  sim_HBRF1S_out[s, ] <- rbinom(K, size = J, prob = mu_phi_HBRF1S * psi_s)
}

qdraw(plot_ppc_degree(obs_degree = d_obs, sim_degree = sim_HBRF1S_out,
                      model_name = "M_HBRF1S(out)",
                      degree_type = "Perturbation out-degree d_k", max_degree = J))
```
